---
title: "Learning R + Hands-On Programming with R"
author: "Richard Cotton & Garret Grolemund"
output:
  pdf_document: default
  html_document: default
---

# R Basics

R operates on vectors.
```{r}
(1:5) # means 1, 2, 3, 4, 5
c(1, 2.5, 3, 4, 4.5) # means 1, 2.5, 3, 4, 4.5

# Non-linear vectors can be enumerated with c()
c(1, 2.5, 3, 4, 4.5) # means 1, 2.5, 3, 4, 4.5


# All arithmetic operators are *vectorized*
(1:5) + c(1, 2.5, 3, 4, 4.5)


# Operations on vectors are ordered
c(1, 3, 4, 4.5, 2.5) == c(1, 2.5, 3, 4, 4.5) # expr contains FALSE

# Many functions are also vectorized (and compute in order)
round(c(1, 2.5, 3, 4, 5))

round(c(1, 3, 5, 4, 2.5))

factorial(c(1, 2.5, 3, 4, 5))

factorial(c(1, 3, 5, 4, 2.5))

round(factorial(c(1, 3, 5, 4, 2.5)))

# ... but agg functions may not be
mean(1:5)

median(c(1, 3, 5, 4, 2.5))

sum(1:5)

sum(c(1, 2.5, 3, 4, 4.5))
```
# Project 1: Weighted Dice

Create an object named `die` that contains the numbers 1 through 6:
```{r}
die <- 1:6
typeof(die)
class(die)
is.vector(die)

# List object names in current R environment
ls()
```

R uses *element-wise execution*. R will apply the same operation to eah element in the set:
```{r}
die + 1

die - 1

die / 1

die / 2

die * die

```

When two or more vectors are used in an operation, R will pair up the vectors and perform the operation on each pair. If the vectors are of unequal lengths, the shorter vector will repeat, starting from it's first element, in order to pair with the remaining elements in the longer vector:
```{r}
die # die is the longer vector
die / c(1, 2)

die # die is the longer vector
die * c(3, 4)

12:24 # [12, 24] is the longer vector
die + 12:24
```

To perform traditional matrix multiplication, do inner multiplication with the `%*%` operator and outer multiplication with the `%o%` operator:
```{r}
die
die %*% die

die
die %o% die
```
Roll the die:
```{r roll}
args(sample) # sample function takes 4 arguments: x, size, replace = FALSE, prob = NULL

sample(die) # sample size default to size of input (x)

sample(die, size = 1 )# size = 1 roll

sample(die, size = 2) # size = 2 rolls
```

Use a data frame to store a deck of cards:
```{r cards}
# First, try a list as a card model
card <- list("ace", "hearts", 1) # list is like a single column from a table
str(card)
typeof(card)
class(card)
is.list(card) # card is a list
is.vector(card) # card is also a vector

# Now create a three card hand with a data.frame
hand <- data.frame(
    face = c("ace",
             "two",
             "six"),       # column 1
    suit = c("clubs",
             "hearts",
             "spades"),    # column 2
    value = c(1,
              2,
              3),          # column 3
    stringsAsFactors = FALSE
)
str(hand)
typeof(hand) # hand is a type of list...
class(hand) # ... of class data.frame
is.data.frame(hand) # hand is a data.frame
is.list(hand) # hand is also a list
is.vector(hand) # hand is *not* a vector

# Now create a deck by reading a csv from the web
deck <- data.frame()
deck_loaded <- tryCatch({
    deck <- read.csv("https://gist.githubusercontent.com/garrettgman/9629323/raw/ee5dfc039fd581cb467cc69c226ea2524913c3d8/deck.csv", sep = ",", na.strings = "NA", strip.white = TRUE, stringsAsFactors = FALSE)

}, error = function (e) {
    print(e) # store error in "message"

}, finally = {})

# print(names(deck_loaded))

if (any(grepl("message", names(deck_loaded)))) {
    print("Load deck from local file")
    deck_file <- "data/deck.csv"
    if (!file.exists("data/deck.csv")) {
        deck_file <- file.choose()
    }
    deck <- read.csv(deck_file, sep = ",", na.strings = "NA", strip.white = TRUE, stringsAsFactors = FALSE)
    str(deck)
}
typeof(deck)
class(deck)
is.data.frame(deck)
```

### R Indices

R has a realy wierd notation for indexing elements. First off, the first element is enumerated by the number 1, instead of 0:
```{r deck}
print(deck)
## 1   king   spades    13
## 2  queen   spades    12
## 3   jack   spades    11
## ...
print(deck[1, c(1, 2, 3)])
## 1   king   spades    13

length(deck) # for each row in deck, return number of *columns*
length(deck[, 1]) # return number of rows in deck

print(deck[1]) # for each row, return a vector cotaining first column and add to data.frame
typeof(deck[1])
class(deck[1]) # data.frame

print(deck[, 1]) # creates single vector, cotaining first column for every row
typeof(deck[, 1])
class(deck[, 1])
is.vector(deck[, 1])

print(deck[1, 1]) # creates vector containing first column of first row
typeof(deck[1, 1])
class(deck[1, 1])
is.vector(deck[1, 1])

print(deck[c(1, 2, 3), 1]) # creates vector containing first column of rows 1, 2, and 3
typeof(deck[c(1, 2, 3), 1])
class(deck[c(1, 2, 3), 1])
is.vector(deck[c(1, 2, 3), 1])

print(deck[c(1:length(deck[,1])), 1]) # creates vector containing first column of revery row
typeof(deck[c(1:length(deck[,1])), 1])
class(deck[c(1:length(deck[,1])), 1])
is.vector(deck[c(1:length(deck[,1])), 1])

print(deck[c(1:length(deck[,1])), c(1:length(deck))]) # for each row, return a vector containing every column of revery row and add to data.frame
typeof(deck[c(1:length(deck[,1])), c(1:length(deck))])
class(deck[c(1:length(deck[,1])), c(1:length(deck))])


```
